---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First lets read the data into a variable while specifying the class of each column:

```{r}
 data <- read.csv("activity.csv",colClasses=c("numeric","Date","numeric"))

```

## What is mean total number of steps taken per day?
In order to answer this question we use a data.table and ggplot2 for plotting a histogram.



```{r global_options,warning=FALSE}
#lets first draw the histogram to get a feel for the data
library(ggplot2)
#then make the histogram
p1 <- qplot(steps,data=data,main="Histogram for number of steps")
p1
```

So it seems that there are quite a few intervals with 0 steps taken.
In order to complement this view, we look at a table showing mean and median for each day.

```{r}

#now on to the mean and median calculation. Load the data.table package
library(data.table)

#then make a data table
dt <- data.table(data)

#then calculate the mean and median per day
meanMedian <- dt[,list(mean=mean(steps,na.rm=TRUE),median=median(steps, na.rm = TRUE)),by=date]
meanMedian
```


Indeed, the median is mostly 0 showing that there are many intervals with 0 steps taken per day.


## What is the average daily activity pattern?
Now we want to look closer at which intervals are active or not.
We fist make a time series plot (line chart) of the 5-minute intervals (x-axis) and the average number of steps taken during each interval, averaged across all days (y-axis)

```{r}
#get the mean steps per interval
meanIntervalSteps <- dt[,list(mean=mean(steps,na.rm=TRUE),median=median(steps, na.rm = TRUE)),by=interval]

#plot them
ggplot(meanIntervalSteps,aes(interval,mean))+geom_line(colour = "royalblue2", size = 1)
```


Now we'll try to calculate Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
```{r}
meanIntervalSteps$interval[which.max(meanIntervalSteps$mean)]
```

So the answer is interval 835. If we assume that the interval value is probably the number of minutes since midnight, then this is maybe this hour of the day (roughly): `r round(835/60)`:00. Maybe the subject takes a walk after lunch?


## Imputing missing values
In order to deal with NAs we'll do the following:

- Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
#the number of NA values is:
length(which(is.na(data$steps)))
```
So the answer is: 2304 NA values

- Next we will replace NA values with the mean for that 5-minute interval over all days and create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
modifiedData <- data

for(i in 1:nrow(modifiedData)){

    if(is.na(modifiedData$steps[i])){
       intval <- modifiedData[i,3]
       #get the average for this day
       avg <- meanIntervalSteps[interval == intval,]
       mean <- avg$mean
       
       #now fill the respective cell
       modifiedData[i,1] <- mean
    }
}

```


- Now we make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. We want to see if these values differ from the estimates from the first part of the assignment. 

```{r}
library(gridExtra)
#first show the new and old histogram side by side to check if we can see a difference
p2 <- qplot(steps,data=modifiedData,main="Histogram for number of steps with NAs replaced by mean per interval")
grid.arrange(p1, p2)
```

The histograms look fairly similar, but there are small difference (see y-axis and the interval around 600 on the x-axis).
Now we'll calculate the differences in mean and median, just to see what the data looks like. If the value is 0 nothing has changed.

```{r}
#make a data table like before
dt2 <- data.table(modifiedData)

#then calculate the mean and median per day
meanMedian2 <- dt2[,list(mean=mean(steps,na.rm=TRUE),median=median(steps, na.rm = TRUE)),by=date]

#check if the values differ from those in the first part of the assignment
differenceTable <- data.frame(meanMedian$mean,meanMedian2$mean,meanMedian$median,meanMedian2$median)
names(differenceTable) <- c("original_mean","new_mean","original_median","new_median")
differenceTable$meanDifference <- differenceTable$original_mean - differenceTable$new_mean
differenceTable$medianDifference <- differenceTable$original_median - differenceTable$new_median
differenceTable
```


We can see that indeed there are a few differences in that mean and median exist for all days now.

- Now, lets answer: What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
#look at the total number of daily steps
#first in the original data
sumSteps <- dt[,list(stepCount=sum(steps,na.rm=TRUE)),by=date]
#now in the replaced data
sumSteps2 <- dt2[,list(stepCount=sum(steps,na.rm=TRUE)),by=date]

differenceTable <- data.frame(sumSteps$stepCount,sumSteps2$stepCount)
names(differenceTable) <- c("stepCountOriginal","stepCountNAReplaced")

differenceTable$Difference <- differenceTable$stepCountOriginal - differenceTable$stepCountNAReplaced
differenceTable
```

We can see that replacing the values influenced those days in which the original step count was 0.


## Are there differences in activity patterns between weekdays and weekends?

- Lets create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
modifiedData$weekDays <- weekdays(modifiedData$date)
modifiedData$weekend <- factor(modifiedData$weekDays)
levels(modifiedData$weekend) <- c("weekday","weekday","weekday","weekday","weekday","weekend","weekend")
```

- Now lets make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
dtweekday <- data.table(subset(modifiedData,weekend=="weekday"))
meanIntervalSteps_weekday <- dtweekday[,list(mean=mean(steps,na.rm=TRUE)),by=interval]

dtweekend <- data.table(subset(modifiedData,weekend=="weekend"))
meanIntervalSteps_weekend <- dtweekend[,list(mean=mean(steps,na.rm=TRUE)),by=interval]


p2 <- ggplot(meanIntervalSteps_weekend,aes(interval,mean))+geom_line(colour = "royalblue2", size = 1) + ggtitle("Weekend")
p1 <- ggplot(meanIntervalSteps_weekday,aes(interval,mean))+geom_line(colour = "royalblue2", size = 1) + ggtitle("Weekday")
grid.arrange(p1, p2)

```

So to answer the question, yes there are differences in patterns. The subject seems to be walking earlier on the weekend than on weekdays and also has more intervals with zero movement on the weekend. Lunchtime seems to involve quite a bit of walking in both cases.
    